Journal and article coverage

### 2025-02-14 analysis/manuscript.tex
First draft data and methods, next results startung with coverage, followed by replication

@@

The aim of this study is to demonstrate the suitability of open
scholarly data sources for assessing the impact of transformative
agreements on hybrid open access. To achieve this, results from
hoaddata, an openly available collection of open research information
regarding hybrid open access, was compared with the established
bibliometric databases Web of Science and Scopus. After describing the
initial data sources used, the necessary pre-processing steps to obtain
eligible articles from transformative agreements using open access
evidence, author roles (first and corresponding) and affiliation data
are presented.

@@

More specifically, hoaddata uses Crossref, a DOI registration agency,
for obtaining journal publication volume and open access status through
Creative Commons licence information relative to the published version
(``version of record''). Because of limited affiliation metadata in
Crossref (Eck \& Waltman, 2022), hoaddata sources first-author
affiliations from OpenAlex.

hoaddata follows good practices for computational reproducibility using
R. The package, which includes data, code, a test suite and
documentation, is openly available on GitHub. To ensure computational
reproducibility while aggregating the data, a GitHub Actions continuous
integration and delivery (CI/CD) workflow interfaces with the SUB
Göttingen's open scholarly data warehouse based on Google BigQuery,
which provides high-performant programmatic access to monthly snapshots
of Crossref and OpenAlex. The workflow has run regularly to fetch
updates from these data sources since 2022. The package version used in
this study is 0.3, containing data from the Crossref 2024-08 dump
provided to Crossref Metadata Plus subscribers and the OpenAlex
2024-08-29 monthly dump. This version including the computation log is
available on GitHub
(\url{https://github.com/subugoe/hoaddata/releases/tag/v.0.3}).

@@

For programmatic access to article-level data, this study used the
database of the Kompetenznetzwerk Bibliometrie (KB) in Germany. The KB
processes raw XML data provided by Clarivate Analytics, which is
provided as an in-house PostgreSQL database under a uniform schema. To
support reproducibility, KB maintains annual snapshots of the database.
Accordingly, this study used the annual snapshot from April 2024,
wos\_b\_202404, which is considered to cover almost the entire previous
publication year (Schmidt et al., 2024).

@@

This study examined the adoption of open access for hybrid journals
included in transformative agreements, which published at least one
journal article in the five-year period 2019-2023. The initial data
source was the Transformative Agreements Public Data from the cOAlition
S Journal Checker Tool (JCT), which helps authors identify suitable open
access publishing venues. Weekly snapshots of this data were archived
using an automated GitHub Actions workflow
(\url{https://github.com/njahn82/jct_data}) and into hoaddata. The
resulting JCT data cover agreements active from July 2021 to July 2024.

The JCT journal data was enriched with the linking ISSN (ISSN-L)
according to the ISSN Registry Agency. A comprehensive exclusion of
fully open access journals was performed, following the methodology
described in (\textbf{jahn\_2025?}). This involved checking multiple
journal lists (DOAJ, OpenAlex, Bielefeld) and article-level
investigations. The JCT institution data was enriched with ROR-IDs from
associated institutions, such as university hospitals or institutes of
large research organisations such as the Max Planck Society, according
to OpenAlex' institution entity.

For each database, article metadata was retrieved using all ISSN
variants linked to an ISSN-L. The metadata included DOIs, document types
(where available), author roles, institutional affiliations, publication
dates and open access information. Publication years were determined
using the earliest known date of publication in a journal. In the case
of Crossref, the article data source for hoaddata, this was the issued
date, although variations were observed (check ref). For Web of Science
and Scopus, the earliest publication date was used where available. In
the latter case, the KB determined the earliest known publication date
by tracking versions of the raw data.

\subsubsection{Open Access}\label{open-access}

Only published versions with Creative Commons licenses available on
publisher platforms were considered as hybrid open access. hoaddata
derived this information from Crossref's license metadata, while Web of
Science and Scopus used data from Unpaywall. In additiaon to Crossref,
Unpaywall also parses publisher websites directly, because some
publishers do not provide machine-readable Creative Commons license
information (Piwowar et al., 2018), although transformative agreement
workflows requiring this information during DOI registration (Geschuhn
\& Stone, 2017).

\subsubsection{Author affiliations}\label{author-affiliations}

Author affiliations at the article level were retrieved for both first
and, if available, corresponding authors across all databases. To handle
different address variants, database-specific affiliation identifiers
were used: ROR-IDs from OpenAlex for hoaddata, affiliation enhanced
names for Web of Science, and Scopus Affiliation Identifiers for Scopus.
Additionally, ISO country codes were retrieved for each author's address
to compile country-level statistics. These country codes for Web of
Science and Scopus were provided by the KB.

Because neither Web of Science nor Scopus support ROR IDs---the
institution identifier used by the JCT---a two-step matching process was
implemented to identify articles covered by transformative agreements.
First, 2,782,540 articles from 6,457 institutions with ROR-IDs in the
JCT data since 2017 (according to hoaddata) were processed to map first
authors' ROR-IDs to corresponding proprietary affiliation identifier in
Web of Science and Scopus using DOI matching. Then, an algorithm
selected the most frequent ROR ID and proprietary identifier pairs to
handle multiple affiliations and organizational hierarchy differences.

@@

@@

As a result of the above-described comprehensive data processing,
data-sets were prepared for each database on open access in hybrid
journals included in transformative agreements at the country and
journal level by year. Table \ref{tab:methods_overview_table} presents
an general overview of the coverage of hybrid journals in transformative
agreements 2019-23 per dataset. It shows that the majority of hybrid
journals published at least one original article or review (core) in the
five-years period. These journals constituted the basis for the
subsequent aggregation of publication metrics. While hoad only covers
articles with DOI, the aggreation of Scopus and Web of Scienc ewas
carried out isng the databse identifier. A subsequent compatision of DOU
coverage illustrates that non.core articles in Web of Scienc e lack
DOIs. These are in aprticular meeting abstracts, a pecularity of this
database. Open access was aggregated using DOI, as all alos Unpaywall
only colelctzs opena cces ststaus information for articles with DOI. A
closer view on core articles with affiliation data reveal a lack of data
withr egard to corresponding auhtors in the case of OpenAlex. Only aroud
two third of investiaged artilces had at affilation data linked to
corresponding authoirs. In case of first authors, the share was higher.
In contrast, web of scienc eand scopus recorded a higher coverage.
Accordingly, only first authro evidence were presented for haoddata in
the follwoing

@@

\section{results}\label{results}

@@

@@

>>>>

% flow chart
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, shapes.multipart}

% Custom styles for the flowchart
\tikzstyle{datasource} = [rectangle, rounded corners, minimum width=2.5cm, minimum height=1cm,text centered, draw=blue!60, fill=blue!5]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=green!60, fill=green!5]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=red!60, fill=red!5]
\tikzstyle{arrow} = [thick,->,>=stealth]
\tikzstyle{output} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm, text centered, draw=purple!60, fill=purple!5]

@@

This study aims to demonstrate the suitability of open scholarly data
sources for assessing the impact of transformative agreements on hybrid
open access. As shown in Figure \ref{fig:workflow}, the methodology
involved comparing hoaddata, an openly available collection of open
research information on hybrid open access, with the bibliometric
databases Web of Science and Scopus. The subsequent section, introduces
the initial data sources followed by a presentation of the necessary
pre-processing steps to obtain eligible articles from transformative
agreements by using author roles (first and corresponding) and
harmonised affiliation data.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[node distance=2cm, thick,scale=1, every node/.style={scale=0.9}]    
% Data Sources
    \node (crossref) [datasource] {Crossref};
    \node (openAlex) [datasource, right=of crossref] {OpenAlex};
    \node (jct) [datasource, right=of openAlex] {JCT};
    \node (wos) [datasource, below=of crossref] {Web of Science};
    \node (scopus) [datasource, below=of jct] {Scopus};

    % Processing Steps
    \node (hoaddata) [datasource, below=1.5cm of $(crossref)!1!(openAlex)$] {hoaddata};
    \node (kb) [process, below=1cm of $(wos)!0.5!(scopus)$] {Journal and article retrieval (including open access)};

 % Matching Process - Fixed line break syntax
    \node (matching) [process, below=1.5cm of $(hoaddata)!1!(kb)$] {Harmonising author affiliations through RoR};

 % Aggregation
    \node (transmatch) [process, below=1.5cm of $(matching)!1!(matching)$] {Estimate TA output};

    % Output
    \node (analysis) [output, below=1.5cm of $(transmatch)!1!(transmatch)$] {Aggregated analytical datasets};

    % Arrows
    \draw [arrow] (crossref) -- (hoaddata);
    \draw [arrow] (openAlex) -- (hoaddata);
    \draw [arrow] (jct) -- (hoaddata);
    \draw [arrow] (hoaddata) -- (kb);
    \draw [arrow] (wos) -- (kb);
    \draw [arrow] (scopus) -- (kb);
    \draw [arrow] (kb) -- (matching);
    \draw [arrow] (matching) -- (transmatch);
    \draw [arrow] (transmatch) -- (analysis);

    % Legend
    \node [datasource, scale=0.7] at ($(analysis)+(4,0)$) {Data Sources};
    \node [process, scale=0.7] at ($(analysis)+(4,-1)$) {Processing Steps};
    \node [output, scale=0.7] at ($(analysis)+(4,-2)$) {Output};
\end{tikzpicture}
\caption{Data processing workflow for comparing hybrid open access uptake across bibliometric data sources. The workflow shows how data from different sources (Crossref, OpenAlex, JCT, Web of Science, and Scopus) is processed and matched using ROR identifiers to enable comparative analysis. TA = Transformative Agreement}
\label{fig:workflow}
\end{figure}

@@

More specifically, hoaddata uses a unified dataset of historic JCT
snapshots since July 2021, which were enriched ISSN variants linked to
an ISSN-L and associated ROR-IDs representing university hospitals or
institutes of large research organisations, according to OpenAlex'
institution entity. A comprehensive exclusion of fully open access
journals using multiple journal lists (DOAJ, OpenAlex, Bielefeld) was
performed to identify hybrid journals, as described in Jahn (2025).
hoaddata relies on Crossref for obtaining journal publication volume and
open access status through Creative Commons licence information relative
to the published version (``version of record''). Because of limited
affiliation metadata in Crossref (Eck \& Waltman, 2022), hoaddata
sources first-author affiliations from OpenAlex.

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow interfaces
with the SUB Göttingen's open scholarly data warehouse based on Google
BigQuery, which provides high-performant programmatic access to monthly
snapshots of Crossref and OpenAlex. The workflow has run regularly to
fetch updates from these data sources since 2022. The package version
used in this study is 0.3, containing data from the Crossref 2024-08
dump provided to Crossref Metadata Plus subscribers and the OpenAlex
2024-08-29 monthly dump. This version including the computation log is
available on GitHub
(\url{https://github.com/subugoe/hoaddata/releases/tag/v.0.3}).

@@

For programmatic access to article-level data, this study used the
database of the Kompetenznetzwerk Bibliometrie (KB) in Germany. The KB
processes raw XML data provided by Clarivate Analytics, which is
provided as an in-house PostgreSQL database under a uniform schema. To
support reproducibility, KB maintains annual snapshots of the database.
Accordingly, this study used the annual snapshot from April 2024
(wos\_b\_202404), which is considered to cover almost the entire
previous publication year (Schmidt et al., 2024).

@@

The hoaddata package (v.0.3) provided an enriched and unified list of
hybrid journals in transformative agreements according to the JCT,
covering agreements active from July 2021 to July 2024 according to the
ESAC registry.

Article metadata was retrieved from each databse using all ISSN variants
linked to an ISSN-L. The metadata included DOIs, publication dates,
article types, open access information as well as author roles and
institutional affiliations.

The publication year for each article was determined using the earliest
known publication date. For articles in hoaddata, this corresponded to
Crossref's issued date.

Publication years were determined using the earliest known date of
publication in a journal. In the case of Crossref, the article data
source for hoaddata, this was the issued date. For Web of Science and
Scopus, the earliest publication date was used where available, with
Scopus dates specifically determined by the KB through version tracking
of the raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews. Because of limited information on these document types in open
scholarly data (Haupka et al., 2024), hoaddata used an extended version
of Unpaywall's paratext recognition approach to exclude non-scholarly
content (Piwowar et al., 2018). To exclude conference supplements, which
are also often not covered by transformative agreements, only articles
published in regular issues, indicated by numerical pagination, were
considered. For Web of Science and Scopus, their established document
type classifications were used to identify original research articles
and reviews, referred to as core articles throughout this study.

\subsubsection{Identifying open access articles in hybrid
journals}\label{identifying-open-access-articles-in-hybrid-journals}

Published articles were considered hybrid open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall. Unpaywall
supplements Crossref metadata by parsing publisher websites directly,
addressing cases where publishers do not provide machine-readable
Creative Commons license information (Piwowar et al., 2018). This
additional parsing remains necessary despite transformative agreement
workflows requiring license information during DOI registration
(Geschuhn \& Stone, 2017). Both Web of Science and Scopus defined hybrid
open access consistently as content available under Creative Commons
licenses on publisher platforms, distinguishing it from bronze open
access that lack such explicit license information.

\subsubsection{Harmonising author affiliations across
databases}\label{harmonising-author-affiliations-across-databases}

Author affiliations were retrieved for both first and, if available,
corresponding authors across all databases to prepare the linking
between articles and institutions covered by transformative agreements.
To handle different address variants, database-specific affiliation
identifiers were used: ROR-IDs from OpenAlex for hoaddata, affiliation
enhanced names for Web of Science, and Scopus Affiliation Identifiers
for Scopus. Additionally, ISO country codes were retrieved for each
author's address to compile country-level statistics. These country
codes for Web of Science and Scopus were provided by the KB.

Because neither Web of Science nor Scopus support ROR IDs---the
institution identifier used by the JCT---a two-step matching process was
implemented to harmonise affiliation data. First, 2,782,540 articles
from 6,457 institutions with ROR-IDs in the JCT data since 2017
(according to hoaddata) were processed to map first authors' ROR-IDs to
corresponding proprietary affiliation identifier in Web of Science and
Scopus using DOI matching. Then, an algorithm selected the most frequent
ROR ID and proprietary identifier pairs to handle multiple affiliations
and organisational hierarchy differences.

@@

Based on these matching tables, articles eligible under transformative
agreements could also be retrieved from Web of Science and Scopus,
although they did not contain the ROR IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions according to
the Transformative Agreement Data dump for each of the data sources
examined. The matching also took into account the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were produced for each database at country and journal level
by year. Table \ref{tab:methods_overview_table} provides a general
overview of the coverage between 2019 and 2023 per database. It shows
that the majority of hybrid journals published at least one original
article or review marked as core during the five-year period. These
journals formed the basis for the subsequent aggregation of publication
metrics.

@@

While hoaddata only covers articles with a DOI, the Scopus and Web of
Science figures were also aggregated using the database identifier. A
subsequent comparison of DOI coverage shows that non-core articles in
Web of Science often lacked a DOI. This was particularly the case for
meeting abstracts, a particular feature of this database. Open access
indicators were aggregated by DOI, as Unpaywall only collects
information on open access status for articles with a DOI. A closer look
at the core articles with affiliation data reveals a lack of data on
corresponding authors in the case of OpenAlex, the affiliation data
source used by hoaddata, compared to Web of Science and Scopus. Only
about two-thirds of the articles examined provided corresponding author
affiliations. For first authors, the proportion was 89\%. Accordingly,
only first author data for hoaddata are considered in the following
analysis.

\section{Results}\label{results}

@@

\bibitem[\citeproctext]{ref-Haupka_2024}
Haupka, N., Culbert, J. H., Schniedermann, A., Jahn, N., \& Mayr, P.
(2024). \emph{Analysis of the publication and document types in
OpenAlex, web of science, scopus, pubmed and semantic scholar}. arXiv.
\url{https://doi.org/10.48550/ARXIV.2406.15154}

@@

\bibitem[\citeproctext]{ref-Marwick_2018}
Marwick, B., Boettiger, C., \& Mullen, L. (2018). Packaging data
analytical work reproducibly using {R} (and friends). \emph{The American
Statistician}, \emph{72}(1), 80--88.
\url{https://doi.org/10.1080/00031305.2017.1375986}
